A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation\tnoteref{ornl-cr}

### 2013-02-19 doc/jcomp2012/jcomp2012.tex
added 9 point convergence explanation in response to jcomp reviewers comments

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five-point and
nine-point stencils were used with both the direct and adjoint
solvers. For each case, $N \times N$ total random walk permutations
were computed per MCSA iteration where $N \times N$ is the number of
discrete grid points in the system. Solver parameters were set to a
weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
computations presented in this section were completed on a 2.66 GHz
Intel Core i7 machine with 8 GB 1067 MHz DDR3
memory. Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed
for each case to converge in seconds, and
Figure~\ref{fig:poisson_iterations} gives the number of iterations
needed for each case to converge relative to the problem size.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A CPU time
    speedup is noted with the adjoint method due to the higher density
    of random walk events in regions with a large residual.}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:MCSA-residual}) was collected at each
iteration for a fixed problem size of
$N=500$. Figure~\ref{fig:poisson_convergence} gives the results of
these computations. First, it is worthy to note on the semilog plot
that we are indeed achieving the expected exponential convergence from
MCSA with both Monte Carlo solvers. Second, we note that using the
adjoint method with the same number of stochastic histores per MCSA
iteration gives a faster rate of converge for the same reasons as
above.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_conv.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=500$. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. A
    higher rate of convergence is observed for MCSA using the adjoint
    Monte Carlo solver as compared to the direct method when both
    solvers compute the same number of random walks per iteration.}
  \label{fig:poisson_convergence}
\end{figure}

\subsection{Comparison to Sequential Monte Carlo}
\label{subsec:sequential comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we
compare its performance to Halton's Sequential Monte Carlo method on
which our previous work in this area was based. For this comparison,
we use the same transient Poisson problem as described in the previous
section and choose only the 5-point stencil to discretize the
Laplacian operator as the previous results yielded little qualitative
difference between the discretizations. Both MCSA and Halton's method
are used with the adjoint Monte Carlo solver. In order to complete the
same study as in the previous section, the number of histories
computed by the Monte Carlo solver at each iteration had to be doubled
to $2 \times N \times N$ in order to ensure convergence in Sequential
Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time
results for this comparison as a function of problem size while
Figure~\ref{fig:seq_iterations} gives the number of iterations to
converge as a function of problem size. In both cases, using the Monte
Carlo solver as a synthetic acceleration rather than in a pure
residual Monte Carlo scheme resulted in a reduction in both CPU time
and iterations required to converge. The additional Richardson
extrapolation between each Monte Carlo solve in the MCSA method gives
a better converged residual source to use with the Monte Carlo
calculation while the Sequential method requires more iterations to
achieve the same level of convergence in the residual.

@@

The benefits of using a synthetic acceleration scheme are also noted
when the infinity norm of the residual computed at each iteration for
both methods was collected at each iteration for a fixed problem sizes
of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
and \ref{fig:seq_500} respectively. In both cases, the Sequential
method is subject to two regimes of exponential convergence with the
later regime converging the slowest while the MCSA method exhibits a
single rate of exponential convergence, much higher than that computed
by Halton's method. Even with the doubling of the number of stochastic
histories computed per time step in order to ensure convergence for
the Sequential method, we still see robustness issues with a
non-monotonically decreasing residual observed for the $N=100$
case. In both cases the MCSA solver is observed to be robust with a
monotonically decreasing residual.

>>>>

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five-point and
nine-point stencils were used with both the direct and adjoint
solvers. For each case, $N \times N$ total random walk permutations
were computed per MCSA iteration where $N \times N$ is the number of
discrete grid points in the system. Solver parameters were set to a
weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance as a function of the problem size. All
computations presented in this section and
\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz
Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A CPU time
    speedup is noted with the adjoint method due to the higher density
    of random walk events in regions with a large residual.}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:MCSA-residual}) was collected at each
iteration for a fixed problem size of
$N=500$. Figure~\ref{fig:poisson_convergence} gives the results of
these computations. First, it is worthy to note on the semilog plot
that we are indeed achieving the expected exponential convergence from
MCSA with both Monte Carlo solvers. Second, we note that using the
adjoint method with the same number of stochastic histores per MCSA
iteration gives a faster rate of converge for the same reasons as
above. We also note here that fewer iterations are required for
convergence when the 9-point stencil is used to discretize the
Laplacian operator (although at no gain in speed as given by the
results in Figure~\ref{fig:poisson_cpu_time}). This is due to the fact
that the smaller discretization error directly corresponds to a more
well defined residual source generated by the Richardson extrapolation
for the Monte Carlo calculation. In addition, the better defined
source is transported through a domain described more accurately by
the 9-point stencil, thus yielding a more accuracte correction vector
from the Monte Carlo calculation.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_conv.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of size $N=500$. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. A
    higher rate of convergence is observed for MCSA using the adjoint
    Monte Carlo solver as compared to the direct method when both
    solvers compute the same number of random walks per iteration.}
  \label{fig:poisson_convergence}
\end{figure}

\subsection{Comparison to Sequential Monte Carlo}
\label{subsec:sequential_comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we
compare its performance to Halton's Sequential Monte Carlo method on
which our previous work in this area was based. For this comparison,
we use the same transient Poisson problem as described in the previous
section and choose only the 5-point stencil to discretize the
Laplacian operator as the previous results yielded little qualitative
difference between the discretizations. Both MCSA and Halton's method
are used with the adjoint Monte Carlo solver. In order to complete the
same study as in the previous section, the number of histories
computed by the Monte Carlo solver at each iteration had to be doubled
to $2 \times N \times N$ in order to ensure convergence in Sequential
Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time
results for this comparison as a function of problem size while
Figure~\ref{fig:seq_iterations} gives the number of iterations to
converge as a function of problem size. In both cases, using the Monte
Carlo solver as a synthetic acceleration rather than in a pure
residual Monte Carlo scheme resulted in a reduction in both CPU time
and iterations required to converge. The additional Richardson
extrapolation between each Monte Carlo solve in the MCSA method gives
a better converged residual source to use with the Monte Carlo
calculation while the Sequential method requires more iterations to
achieve the same level of convergence in the residual.

@@

The benefits of using a synthetic acceleration scheme are also noted
when the infinity norm of the residual computed at each iteration for
both methods was collected at each iteration for a fixed problem sizes
of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
and \ref{fig:seq_500} respectively. In both cases, the Sequential
method is subject to two regimes of exponential convergence with the
later regime converging the slowest while the MCSA method exhibits a
single rate of exponential convergence observed to be much higher than
that computed by Halton's method. Even with the doubling of the number
of stochastic histories computed per time step in order to ensure
convergence for the Sequential method, we still see robustness issues
with a non-monotonically decreasing residual observed for the $N=100$
case. In both cases the MCSA solver is observed to be robust with a
monotonically decreasing residual.
